Tropical Mathematics and the Lambda Calculus

### 2023-01-22 section6.tex
commit

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}. 
In this section we first reconstruct this correspondence, by combining some {folklore} results with more abstract ones from recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}. Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the exponential comonad of $\LREL$.

@@

\subsection{$\Lawv$-Modules}

@@

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a SMCC.
\end{proposition}
%\begin{proof}
The hom-sets $\Mod(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
$\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
Finally, both products and coproducts in $\Mod$ are given by the Cartesian products of the underlying posets, with action defined pointwise.
%
%, similarly to the case of sup-lattices \cite{}, as the quotient of the free sup-lattice $\C P(M\times N)$ under the smallest congruence containing all pairs $((\bigvee A,y),\bigcup_{a\in A}\{(a,y)\})$, for $A\subseteq M, y\in N$, 
%$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
%\end{proof}

@@

\subsection{$\Lawv$-Categories}

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}.%, or as a \emph{quasi-metric space}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.

@@

For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
Notice that, when $X$ has the discrete metric, the metric space $[X,\Lawv]^{\sym}$ coincides with $\Lawv^{X}$ with the norm metric.
The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.

Actually, a well-studied example of $\Lawv$-categories is provided by $\Lawv$-modules.

@@

\subsection{Complete $\Lawv$-Categories Correspond to $\Lawv$-Modules}

@@

Since $\Mod$ (and thus $\GMet$ too) is a SMCC, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,Y$, we have that the distance on the hom-set $\Mod(X,Y)$ is given by 
\begin{align}
\Mod(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
\end{align}
and the distance on basic tensors 
of $X\otimes Y$ is given by
\begin{align}
%(X\otimes Y)(\alpha, \beta)=
%\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
(X\otimes Y)(x\otimes y, x'\otimes y')=
X(x,x')+Y(y,y')
\end{align}
and is extended to arbitrary joins by continuity.
%where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
%and thus coincides with the sum-metric over basic tensors;
%\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
%\begin{align}
%(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
%\end{align}

@@

\subsection{Exponential and Differential Structure}

>>>>

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}. 
In this section we first reconstruct this correspondence, by combining some {folklore} results with more abstract ones from recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}. Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the Lafont exponential of $\LREL$.

@@

\subsection{$\Lawv$-modules}

@@

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a SMCC.
\end{proposition}
%\begin{proof}
The hom-sets $\Mod(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. The tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
$\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
Finally, $\Mod$ has \emph{biproducts}: products and coproducts are both given by the Cartesian product of the underlying posets, with action defined pointwise.
%
%, similarly to the case of sup-lattices \cite{}, as the quotient of the free sup-lattice $\C P(M\times N)$ under the smallest congruence containing all pairs $((\bigvee A,y),\bigcup_{a\in A}\{(a,y)\})$, for $A\subseteq M, y\in N$, 
%$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
%\end{proof}

@@

\subsection{$\Lawv$-categories}

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$ satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred to as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}. %, or as a \emph{quasi-metric space}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, as functoriality reads as $Y(f(x),f(y))\leq X(x,y)$.

@@

For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
Notice that, when $X$ has the discrete metric, the metric space $[X,\Lawv]^{\sym}$ coincides with $\Lawv^{X}$ with the $\infty$-norm metric.
The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.

Actually, an important example of $\Lawv$-categories are precisely the $\Lawv$-modules:

@@

\subsection{Complete $\Lawv$-categories correspond to $\Lawv$-modules}

@@

Since $\Mod$ (and thus $\GMet$ too) is a SMCC, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,Y$, we have that the distance on the hom-set $\Mod(X,Y)$ is given by 
\begin{align}
\Mod(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
\end{align}
and the distance on $X\otimes Y$ is given by
\begin{align}
(X\otimes Y)(\alpha, \beta)=
\sup_{i}\inf_{j}\left\{X(x_{i},x'_{j})+Y(y_{j},y'_{j}\right\},
\end{align}
 where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
from which we deduce that, for basic tensors
$x\otimes y, x'\otimes y'$, their distance is just $X(x,x')+Y(y,y')$.

%
%
%and is extended to arbitrary joins by continuity, i.e.~
%$(X\otimes Y)(\alpha, \beta)=
%\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
%$, where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$.
%%and thus coincides with the sum-metric over basic tensors;
%\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
%\begin{align}
%(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
%\end{align}

@@

\subsection{Exponential and differential structure}

### 2023-01-22 section8.tex
commit

Finally, as suggested in Section \ref{section5}, an interesting direction to look at is the semantics of finiteness spaces \cite{Ehrhard2005}, which allows to model typable terms by means of \emph{finitary} Taylor expansions. Indeed, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications (hence by a Lipschitz function in our setting), may allow one to predict how sensitive $M$ will be ``around $N$''.
%
%We have shown that:
%
%- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)
%
%- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)
%
%- this relation \emph{seems} to provide applications in different fields (section 5).
%
%Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
%For instance, the general questions are:
%
%- Can we improve the results of section 4 ?
%
%- Can we develop and make the applications of section 5 useful ?
%
%- What does the general setting of section 6 give in terms of theoretical and applied results ?
%
%- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)
%
%- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
%Say what could they do and why.

>>>>

Finally, since bounds on the Taylor expansion translate into Lipschitz conditions, 
 two interesting directions to explore are provided by (non-idempotent) intersection types and finiteness spaces \cite{Ehrhard2005}, as both methods are in principle capable of capturing \emph{finitary} bounds on the Taylor expansion. %
Notably, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications may allow one to predict how sensitive $M$ will be ``around $N$''. 
%We have shown that:
%
%- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)
%
%- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)
%
%- this relation \emph{seems} to provide applications in different fields (section 5).
%
%Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
%For instance, the general questions are:
%
%- Can we improve the results of section 4 ?
%
%- Can we develop and make the applications of section 5 useful ?
%
%- What does the general setting of section 6 give in terms of theoretical and applied results ?
%
%- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)
%
%- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
%Say what could they do and why.

### 2023-01-22 tropicalization.tex
commit

%For any positive real $t$, the tropical polynomials are in one-to-one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial 
%$f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
%%Hence, for any polynomial $p(x)= 
%%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
%%\begin{align}
%%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
%%\end{align}
%More generally, 
Let us fix a positive real $t>0$. For any function $f:[0,1]\to [0,1]$ which can be written as a parameterized {power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$, 
% (as we'll see in Section \ref{section5}, such functions arise naturally from the interpretation of probabilistic programs),
  we let its \emph{tropicalization} be the tLs $\trop f: \Lawv \to \Lawv$ defined as follows:
\begin{align}\label{eq:defTLS}
\trop f(\alpha) =\inf_{n}\left\{ n\alpha+c_{n}\right\}
\end{align}
%Such functions, called \emph{tropical Laurent series} \cite{Porzio2021}, will be discussed in more detail in Section \ref{section5}.
Clearly, for any $t>0$, there is a one-to-one correspondence between the representation of power series in parameterized form and the associated tLs. Moreover, 
$f$ and $\trop f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
$\trop_{t}f: [0,\infty]\to [0,\infty]$ be defined by 
$\trop_{t}f(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
$\trop f= \lim_{t\to 0}\trop_{t}f$. 
Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
$\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$.

>>>>

%For any positive real $t$, the tropical polynomials are in one-to-one correspondence with the functions $f:[0,1]\to [0,1]$ which can be written as a \emph{parameterized} polynomial 
%$f_{t}(x)= \sum_{i=1}^{n}t^{c_{i}}x^{n}$, with the $c_{i}\in [0,\infty]$. 
%%Hence, for any polynomial $p(x)= 
%%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
%%\begin{align}
%%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
%%\end{align}
%More generally, 
Let us fix a positive real $t>0$. For any function $f:[0,1]\to [0,1]$ which can be written as a parameterized {power series} of the form $f_{t}(x)= \sum_{n}t^{c_{n}}x^{n}$, 
% (as we'll see in Section \ref{section5}, such functions arise naturally from the interpretation of probabilistic programs),
  we let its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ be the tLs defined as follows:
\begin{align}\label{eq:defTLS}
\trop f(\alpha) =\inf_{n}\left\{ n\alpha+c_{n}\right\}
\end{align}
%Such functions, called \emph{tropical Laurent series} \cite{Porzio2021}, will be discussed in more detail in Section \ref{section5}.
Clearly, for any $t>0$, there is a one-to-one correspondence between the representations of power series in parameterized form and the associated tLs. Moreover, 
$f$ and $\trop f$ can be related by a limit passage as follows: the functions $\phi_{t}(x)= -\log_{t}x$ and $\varphi_{t}(\alpha)= t^{-\alpha}$ define continuous bijections between $[0,1]$ and $[0,\infty]$ and, by letting
$\trop_{t}f: [0,\infty]\to [0,\infty]$ be defined by 
$\trop_{t}f(\alpha)= \phi_{t}\circ f \circ \psi_{t}$, one has that 
$\trop f= \lim_{t\to 0}\trop_{t}f$. 
Indeed, one can check that the ``parameterized'' sums and product $\alpha \sumt{t}\beta:= \phi_{t}(\psi_{t}(\alpha)+\psi_{t}(\beta))= \log_{t}(t^{-\alpha}+t^{-\beta})$ and 
$\alpha \prodt{t}\beta:= \phi_{t}( \psi_{t}(\alpha)\psi_{t}(\beta))=
\log_{t}(t^{-\alpha}t^{-\beta})$ converge respectively to $\min\{\alpha,\beta\}$ and $\alpha+\beta$, when $t\to 0$.
